---
title: "Untitled"
format: html
editor: visual
---

## Quarto

### Load / Setup

```{r}
#load libraries
library(tidycensus)
library(tidyverse)

#set API key
apiKey="03fee447fb8eb2316ca7a1579efac85a6e5050de"
census_api_key(apiKey, install = TRUE, overwrite=TRUE)
readRenviron("~/.Renviron")
```

## Querying

```{r}
# Make 2023 dataframe to view all variables
vars_2023 <- load_variables(2023, "acs5", cache = TRUE)
```

### Find Variables

B24114 --\> "Detailed Occupation for the Civilian Employed Population 16 Y..." C24010 --\> "Sex by Occupation for the Civilian Employed Population 16..." C24030 --\> "Sex by industry" B24030 --\> Not in data, is uncollapsed version of C24030

```{r}
# Create dataframes for neccessary table(s)

C24030_vars <- vars_2023 %>%
  filter(grepl("^C24030_", name))

```

There are 26 variables but 13 are subcategories, so we only need the major 13. We will total to 26 variables due to the seperation by sex.

Male: C24030_003 - Agriculture, forestry, fishing and hunting, and mining C24030_006 - Construction C24030_007 - Manufacturing C24030_008 - Wholesale trade C24030_009 - Retail trade C24030_010 - Transportation and warehousing, and utilities C24030_013 - Information C24030_014 - Finance and insurance, and real estate and rental and leasing C24030_017 - Professional, scientific, and management, and administrative and waste management services C24030_021 - Educational services, and health care and social assistance C24030_024 - Arts, entertainment, and recreation, and accommodation and food services C24030_027 - Other services, except public administration C24030_028 - Public Administration

Female: C24030_030 - Agriculture, forestry, fishing and hunting, and mining C24030_033 - Construction C24030_034 - Manufacturing C24030_035 - Wholesale trade C24030_036 - Retail trade C24030_037 - Transportation and warehousing, and utilities C24030_040 - Information C24030_041 - Finance and insurance, and real estate and rental and leasing C24030_044 - Professional, scientific, and management, and administrative and waste management services C24030_048 - Educational services, and health care and social assistance C24030_051 - Arts, entertainment, and recreation, and accommodation and food services C24030_054 - Other services, except public administration C24030_055 - Public Administration

### Query 2023 Data

```{r}

industry2023 <- get_acs(
  geography = "county", # could be tract, zip
  variables = c(
    "M_Agriculture, forestry, fishing and hunting, and mining" = "C24030_003",
    "M_Construction" = "C24030_006",
    "M_Manufacturing" = "C24030_007",
    "M_Wholesale trade" = "C24030_008",
    "M_Retail trade" = "C24030_009",
    "M_Transportation and warehousing, and utilities" = "C24030_010",
    "M_Information" = "C24030_013",
    "M_Finance and insurance, and real estate and rental and leasing" = "C24030_014",
    "M_Professional, scientific, and management, and administrative and waste management services" = "C24030_017",
    "M_Educational services, and health care and social assistance" = "C24030_021",
    "M_Arts, entertainment, and recreation, and accommodation and food services" = "C24030_024",
    "M_Other services, except public administration" = "C24030_027",
    "M_Public Administration" = "C24030_028",
    "F_Agriculture, forestry, fishing and hunting, and mining" = "C24030_030",
    "F_Construction" = "C24030_033",
    "F_Manufacturing" = "C24030_034",
    "F_Wholesale trade" = "C24030_035",
    "F_Retail trade" = "C24030_036",
    "F_Transportation and warehousing, and utilities" = "C24030_037",
    "F_Information" = "C24030_040",
    "F_Finance and insurance, and real estate and rental and leasing" = "C24030_041",
    "F_Professional, scientific, and management, and administrative and waste management services" = "C24030_044",
    "F_Educational services, and health care and social assistance" = "C24030_048",
    "F_Arts, entertainment, and recreation, and accommodation and food services" = "C24030_051",
    "F_Other services, except public administration" = "C24030_054",
    "F_Public Administration" = "C24030_055"
  ),
  state = "OR",
  county = "Multnomah",
  year = 2023,
  survey = "acs5"
)

```

### 10 Year Query

```{r}
# Define Years
years <- 2013:2023

# Create Empty List to Store Results
pop_data_list <- list()

# Loop for Population Data
for (year in years) {
  pop_data_list[[as.character(year)]] <- get_acs(
    geography = "county",
    variables = c(
      "M_Agriculture, forestry, fishing and hunting, and mining" = "C24030_003",
      "M_Construction" = "C24030_006",
      "M_Manufacturing" = "C24030_007",
      "M_Wholesale trade" = "C24030_008",
      "M_Retail trade" = "C24030_009",
      "M_Transportation and warehousing, and utilities" = "C24030_010",
      "M_Information" = "C24030_013",
      "M_Finance and insurance, and real estate and rental and leasing" = "C24030_014",
      "M_Professional, scientific, and management, and administrative and waste management services" = "C24030_017",
      "M_Educational services, and health care and social assistance" = "C24030_021",
      "M_Arts, entertainment, and recreation, and accommodation and food services" = "C24030_024",
      "M_Other services, except public administration" = "C24030_027",
      "M_Public Administration" = "C24030_028",
      "F_Agriculture, forestry, fishing and hunting, and mining" = "C24030_030",
      "F_Construction" = "C24030_033",
      "F_Manufacturing" = "C24030_034",
      "F_Wholesale trade" = "C24030_035",
      "F_Retail trade" = "C24030_036",
      "F_Transportation and warehousing, and utilities" = "C24030_037",
      "F_Information" = "C24030_040",
      "F_Finance and insurance, and real estate and rental and leasing" = "C24030_041",
      "F_Professional, scientific, and management, and administrative and waste management services" = "C24030_044",
      "F_Educational services, and health care and social assistance" = "C24030_048",
      "F_Arts, entertainment, and recreation, and accommodation and food services" = "C24030_051",
      "F_Other services, except public administration" = "C24030_054",
      "F_Public Administration" = "C24030_055"
      ),
    state = "OR",
    county = "Multnomah",
    year = year,
    survey = "acs5",
    geometry = FALSE #we'll look at TRUE next week
  ) |>
    mutate(year = year) # add year column for easier joining later
}

# Combine All Years
industry10YR <- bind_rows(pop_data_list)

```

## Wrangling

### 10YR Query

```{r}
# Fix Sex Division

fixed_10YR <- industry10YR |> 
  mutate(industrycategory = str_extract(variable, "(?<=\\_).*")) |> 
  mutate(industrycategory = str_to_lower(industrycategory)) |> 
  group_by(GEOID, NAME, year, industrycategory) |> 
  summarize(population = sum(estimate, na.rm = TRUE),
            moe = sqrt(sum(moe^2)),  
            .groups = "drop")
  
```

### Crosswalk File

```{r}
# Download Condensed Crosswalk File (done in google sheets)
industrycodes <- read_csv("2017industrycodelist.csv")

# Rename Crosswalk 
industrycodes <- industrycodes |>
  rename(industrycategory = "Major Category") |> 
  rename(naicscode = "2017 NAICS Code") |> 
  rename(censuscode = "2017 Census Code")

# Fix Crosswalk (turn into applyable ranges)
industrycodes <- industrycodes |> 
  separate(naicscode, into = c("naicscodestart", "naicscodeend"), sep = "-", convert = TRUE) |> 
  mutate(naicscodeend = if_else(is.na(naicscodeend), naicscodestart, naicscodeend)) |> 
  separate(censuscode, into = c("censusstart", "censusend"), sep = "-", convert = FALSE ) |> 
  mutate(censusend = if_else(is.na(censusend), censusstart, censusend),
         censusend = as.integer(censusend),
         censusstart = as.integer(censusstart))

industrycodes[1, "industrycategory"] <- "Agriculture, Forestry, Fishing and Hunting, and Mining"
```

### Aggregated Data

```{r}
# Download Aggregated Data from mayoblt
naics_agg <- read_csv("naics_aggregated.csv")
census_agg <- read_csv("census_aggregated.csv")

# Use crosswalk to prepare aggregated data

# NAICS Crosswalk
naics_agg2 <- left_join(naics_agg, industrycodes, join_by(naicscode >= naicscodestart,
                                                          naicscode <= naicscodeend))

naics_aggregated <- naics_agg2 |>
  mutate(industrycategory = str_to_lower(industrycategory)) |> 
  group_by(ddodyear, industrycategory) |> 
  summarize(deathcount = sum(deathcount),
            .groups = "drop")

# CENSUS Crosswalk
census_agg <- census_agg |> 
  mutate(censuscode = as.integer(censuscode))

census_agg2 <- left_join(census_agg, industrycodes, join_by(censuscode >= censusstart,
                                                            censuscode <= censusend))

census_aggregated <- census_agg2 |> 
  mutate(industrycategory = str_to_lower(industrycategory)) |> 
  group_by(ddodyear, industrycategory) |> 
  summarize(deathcount = sum(deathcount),
            .groups = "drop")


# Check if Codes are Same
# all.equal(census_aggregated, naics_aggregated)

```

## Combine

### Query + NAICS-Ind

```{r}
joined_NAICS <- 
  left_join(fixed_10YR, naics_aggregated, by = c("industrycategory", "year" = "ddodyear"))

# Create Rate Column
joined_NAICS <- joined_NAICS |> 
  mutate(rate_per_100k = (deathcount / population) * 100000)
```

### Query + CensusCodes-Ind

```{r}
joined_CENSUS <- 
  left_join(fixed_10YR, census_aggregated, by = c("industrycategory", "year" = "ddodyear"))

# Create Rate Column
joined_CENSUS <- joined_CENSUS |> 
  mutate(rate_per_100k = (deathcount / population) * 100000)
```

### Rename Categories for Visualizations

```{r}
industryrenaming <- read_csv("2017crosswalk_renaming.csv")
industryrenaming[1, "industrycategory_old"] <- "Agriculture, Forestry, Fishing and Hunting, and Mining"
industryrenaming <- industryrenaming |> mutate(industrycategory_old = str_to_lower(industrycategory_old))

joined_NAICS <- left_join(joined_NAICS, industryrenaming, by = c("industrycategory" = "industrycategory_old"))
joined_NAICS <- joined_NAICS |> 
  select(-industrycategory) |> 
  rename(industrycategory = industrycategory_new) |> 
  relocate(industrycategory, .after = year)

# Create Averages Dataframe

average_naics <- joined_NAICS |> 
  group_by(industrycategory) |> 
  summarize(deathrate = mean(rate_per_100k), sd = sd(rate_per_100k))

```

## Visualizations

### Testing

#### Industry Codes

```{r}
joined_NAICS |> 
  ggplot(aes(year, rate_per_100k, color = industrycategory)) +
  geom_line() 

joined_CENSUS |> 
  mutate(industrycategory = word(industrycategory)) |> 
  ggplot(aes(year, rate_per_100k, color = industrycategory)) +
  geom_line() 
```

```{r}
joined_NAICS |> 
  mutate(industrycategory = word(industrycategory)) |> 
  ggplot(aes(industrycategory, rate_per_100k, fill = industrycategory)) +
  geom_boxplot() +
  labs(title = "boxplot by industry")

joined_NAICS |> 
  mutate(industrycategory = word(industrycategory)) |> 
  ggplot(aes(fct_reorder(industrycategory, rate_per_100k), rate_per_100k, fill = industrycategory)) +
  geom_boxplot() +
  coord_flip() +
  labs(title = "boxplot by industry")

joined_NAICS |> 
  mutate(industrycategory = word(industrycategory)) |> 
  ggplot(aes(industrycategory, rate_per_100k, fill = industrycategory)) +
  geom_col() +
  facet_wrap(vars(year)) +
  labs(title = "geom col faceted by year")
```

```{r}
joined_NAICS |> 
  mutate(industrycategory = word(industrycategory)) |>
  #filter(industrycategory %in% t3b3) |> 
  ggplot(aes(industrycategory, rate_per_100k, fill = industrycategory)) +
  geom_violin()
```

#### Averaged Data

```{r}
avg_NAICS <- joined_NAICS |> 
  mutate(industrycategory = word(industrycategory)) |> 
  group_by(industrycategory) |> 
  summarize(deathrate = mean(rate_per_100k), sd = sd(rate_per_100k))

avg_NAICS |> 
  ggplot(aes(fct_reorder(industrycategory,deathrate), deathrate, fill = industrycategory)) +
  geom_col() +
  coord_flip() +
  labs(title = "Average Rate per Category from all Years (2013-2023)")

avg_NAICS |> 
  ggplot(aes(fct_reorder(industrycategory,deathrate), deathrate, fill = industrycategory)) +
  geom_col() +
  geom_errorbar(aes(ymin = deathrate-sd, 
                    ymax = deathrate+sd),
                width=0.5) +
  #coord_flip() +
  labs(title = "Previous Plot w/ Error Bars")

avg_NAICS |> 
  ggplot(aes(fct_reorder(industrycategory,deathrate), deathrate, fill = industrycategory)) +
  geom_point() +
  geom_hline(aes(yintercept = mean(deathrate), color = "red"))
  #coord_flip()

```

### Real / Finished

```{r}
p1 <- joined_NAICS |> 
  ggplot(aes(year, rate_per_100k, color = industrycategory)) +
  geom_line() +
  scale_x_continuous(breaks = c(2014,2016,2018,2020,2022)) +
  labs(title = "CV Death Rate by Industry Category",
       subtitle = "Multnomah County, 2013-2023",
       y = "CV Death Rate (per 100k)",
       x = "Year",
       color = "Industry")

p2 <- joined_NAICS |> 
  ggplot(aes(fct_reorder(industrycategory, rate_per_100k), rate_per_100k)) +
  geom_boxplot() +
  coord_flip() +
  labs(title = "Box Plot of CV Death Rates by Occupation",
       subtitle = "Mulnomah County, 2013-2023",
       y = "CV Death Rate (per 100k)",
       x = "Industry Category")

p3 <- average_naics |> 
  ggplot(aes(fct_reorder(industrycategory,deathrate), deathrate, fill = deathrate)) +
  geom_col() +
  scale_fill_gradient(low = "#00C08D", high = "#00ABFD") +
  geom_errorbar(aes(ymin = deathrate-sd, 
                    ymax = deathrate+sd),
                width=0.5) +
  labs(title = "CV Death Rate by Occupation",
       subtitle = "Multnomah County, Average Counts from 2013-2023",
       y = "Average Death Rate from 2013-2023 (per 100k)",
       x = "") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Save All Plots

ggsave("FinalVis/linechart.png", plot = p1, width = 10, height = 6)
ggsave("FinalVis/boxplot.png", plot = p2)
ggsave("FinalVis/colchart.png", plot = p3)
```

## Statistics

```{r}
# Take top 3 and bottom 3

t3b3 <- c("wholesale", "professional,", "arts,", "construction", "transportation", "agriculture,")

dr_maxmin <- joined_NAICS |> 
  mutate(industrycategory = word(industrycategory)) |> 
  filter(industrycategory %in% t3b3)

# ANOVA

model <- aov(rate_per_100k ~ industrycategory,
             data = dr_maxmin,
             weights = population)
summary(model)


# TukeyHSD

tukey_model <- TukeyHSD(model)

tukey_frame <- as.data.frame(tukey_model$industrycategory) 

tukey_frame <- tukey_frame |> 
  select(-lwr, -upr) |> 
  mutate(Comparison = rownames(tukey_frame), .before = diff) |> 
  rename(Difference = diff,
         "P-Value" = "p adj")

rownames(tukey_frame) <- (1:15)

tukey_frame
```
